Quantizing the Bosonic String on a Loop Quantum Gravity Background

### 2024-06-22 lqg-strings.tex
Fixed one reference, other minor changes.

In \autoref{sec:emergent-dimensions} we complement Afshordi's work \cite{Afshordi-emergent} with insights from LQG, showing that LQG quantum geometry naturally constrains most of the free parameters of said model, leading to a more concrete theory. We also suggest a possible consequence for the parameters that remain unconstrained and a generalization of the model.

@@

There is another alternative to incorporate LQG quantum-geometric effects into the string action and get something conceptually similar to what we had in section \ref{sec:inverse-metric-Polyakov}: Afshordi's \textquotedblleft emergent dimension" \cite{Afshordi-emergent} model, with the action being
\begin{multline} \label{eqn:emergent-dimensions-action}
	S_{ED} = - \frac{1}{4 \pi \alpha '} \int \diff ^2 x \sqrt{- g} \big( g^{a b} \del _{a} X^{\mu} \del _{b} X^{\nu} \eta _{\mu \nu} + \\
	+ \mu _{0} ^2 \e ^{- m_{\mu} / T} X^{\mu} X^{\nu} \eta _{\mu \nu} \big)
\end{multline}
where $T$ is the \emph{temperature} of target space and the parameters $m_{\mu}$ satisfy an ordering relation $m_{0} \gg m_{1} \gg ... \gg m_{d}$. This is Afshordi's model \cite{Afshordi-emergent}. Where our approach diverges is that we realize the parameters $m_{\mu}$ are constrained by the modified dispersion relations of each field $X^{\mu}$, since as \cite{Bodendorfer-LQG} points out, LQG quantum geometry breaks Lorentz invariance near the Planck scale through modifying the dispersion relation by constraining the free parameters of the model as
\begin{equation} \label{eqn:LQG-modified-dispersion-relation}
	E^2 = m^2 + p^2 + \sum _{n \geq 3} c_{n} \frac{p^n}{E_{pl} ^{n - 2}} ,
\end{equation}
so we note that the mass parameters of the fields $X^{\mu}$ will be constrained as
\begin{equation} \label{eqn:X-masses-constraint}
	E_{\alpha} ^2 = p_{\alpha} ^2 + \sum _{n = 3} ^{D - 1} \frac{m_{n}}{T} \frac{p_{\alpha} ^{n}}{E_{pl} ^{n - 2}} ,
\end{equation}
where $E_{\alpha} ^2$ does not mean $E_{\alpha} E^{\alpha}$, rather it means $(E_{3})^2 , (E_{4})^2 , ... , (E_{D - 1})^2$, \ie, the squared WS energy (and momentum) of each field $X^{\mu}$.

@@

Therefore, now we will require that the bulk connection and the bulk vielbein can be pulled back onto a connection $\mc{A}_{a}^{ij}$ and a dyad $e^a_i$, respectively, living on the string worldsheet $\Sigma$. $X^\mu(\tau , \sigma)$ are the space-time coordinates of a point $(\tau , \sigma)$ on the worldsheet. Now if one adopts the ``sigma model'' point of view then the $X$'s can be viewed as $d+1$ scalar fields living on a $1+1$ worldsheet. But we know that these fields are not a collection of $d+1$ independent scalars. They are components of a $d+1$ dim vector since they transform with the fundamental representation of $\mathrm{SO}(d,1)$ under global Lorentz transformations acting on the ambient space-time. However, the symmetries of a flat background include not only Lorentz transformations bulk also translations. The string action is required to be invariant under both of these:
\begin{align*}
	X^\mu & \rarrow X^\mu + \delta X^\mu; \quad \delta X^\mu \in \mbb{R}^{3,1}            \\
	X^\mu & \rarrow \Lambda^\mu{}_\nu X^\mu; \quad \Lambda^\mu{}_\nu \in \mathrm{SO}(3,1)
\end{align*}
where $\delta X^\mu$ is some arbitrary vector in Minkowski space. The semidirect product of $\mbb{R}^{3,1}$ with $\mathrm{SO}(3,1)$ is $\mbb{P}^{3,1}$, the Poincaré group in $3+1$ dimensions. Of course, these are good symmetries only if our background is flat. An arbitrarily curved background can have one or more Killing vectors depending on how much symmetry the geometry has, but \emph{in general} even that need not be the case. The best one can hope for is some notion of local flatness is some finite subregion of the background manifold. Moreoever the entire purpose of this work to try and understand how to write down a string action in an \emph{arbitrarily curved} background, assuming that such a thing is even possible.

Even in an arbitrarily curved background, however, symmetry still plays a very important role. Even if we have discarded \emph{global} symmetries, that still leaves us with \emph{local} symmetries in the form of diffeomorphisms under which must the action must remain invariant. If, in addition, we have fields such as a gauge connection and an associated vielbein living in the background, then we also have to take local gauge transformations of these objects into account.

However, diffeomorphisms and gauge transformations are not independent of each other. A diffeomorphism is a differentiable map from one manifold to another (in the case of general relativity both manifolds are generally taken to be the same), which has a differentiable (and invertible) map. In more pedestrian language, a diffeomorphism is an (invertible) change of co-ordinates on a manifold. A gauge transformation is a redefinition of the local fields which live on the underlying manifold. In fact, these two can be seen to be equivalent to each other, with diffeomorphisms representing ``passive'' co-ordinate transformations and gauge representing ``active'' co-ordinate transformations. Therefore, given any spacetime the only question is \emph{which} gauge group is required to capture the full information about diffeomorphisms of that spacetime.

>>>>

In \autoref{sec:emergent-dimensions} we complement Afshordi's work \cite{Afshordi2014emergent} with insights from LQG, showing that LQG quantum geometry naturally constrains most of the free parameters of said model, leading to a more concrete theory. We also suggest a possible consequence for the parameters that remain unconstrained and a generalization of the model.

@@

There is another alternative to incorporate LQG quantum-geometric effects into the string action and get something conceptually similar to what we had in section \ref{sec:inverse-metric-Polyakov}: Afshordi's \textquotedblleft emergent dimension" \cite{Afshordi2014emergent} model, with the action being
\begin{multline} \label{eqn:emergent-dimensions-action}
	S_{ED} = - \frac{1}{4 \pi \alpha '} \int \diff ^2 x \sqrt{- g} \big( g^{a b} \del _{a} X^{\mu} \del _{b} X^{\nu} \eta _{\mu \nu} + \\
	+ \mu _{0} ^2 \e ^{- m_{\mu} / T} X^{\mu} X^{\nu} \eta _{\mu \nu} \big)
\end{multline}
where $T$ is the \emph{temperature} of target space and the parameters $m_{\mu}$ satisfy an ordering relation $m_{0} \gg m_{1} \gg ... \gg m_{d}$. This is Afshordi's model \cite{Afshordi2014emergent}. Where our approach diverges is that we realize the parameters $m_{\mu}$ are constrained by the modified dispersion relations of each field $X^{\mu}$, since as \cite{Bodendorfer-LQG} points out, LQG quantum geometry breaks Lorentz invariance near the Planck scale through modifying the dispersion relation by constraining the free parameters of the model as
\begin{equation} \label{eqn:LQG-modified-dispersion-relation}
	E^2 = m^2 + p^2 + \sum _{n \geq 3} c_{n} \frac{p^n}{E_{pl} ^{n - 2}} ,
\end{equation}
so we note that the mass parameters of the fields $X^{\mu}$ will be constrained as
\begin{equation} \label{eqn:X-masses-constraint}
	E_{\alpha} ^2 = p_{\alpha} ^2 + \sum _{n = 3} ^{D - 1} \frac{m_{n}}{T} \frac{p_{\alpha} ^{n}}{E_{pl} ^{n - 2}} ,
\end{equation}
where $E_{\alpha} ^2$ does not mean $E_{\alpha} E^{\alpha}$, rather it means $(E_{3})^2 , (E_{4})^2 , ... , (E_{D - 1})^2$, \ie, the squared WS energy (and momentum) of each field $X^{\mu}$.

@@

Therefore, now we will require that the bulk connection and the bulk vielbein can be pulled back onto a connection $\mc{A}_{a}^{ij}$ and a dyad $e^a_i$, respectively, living on the string worldsheet $\Sigma$. $X^\mu(\tau , \sigma)$ are the space-time coordinates of a point $(\tau , \sigma)$ on the worldsheet. Now if one adopts the ``sigma model'' point of view then the $X$'s can be viewed as $d+1$ scalar fields living on a $1+1$ worldsheet. But we know that these fields are not a collection of $d+1$ independent scalars. They are components of a $d+1$ dim vector since they transform with the fundamental representation of $\mathrm{SO}(d,1)$ under global Lorentz transformations acting on the ambient space-time. However, the symmetries of a flat background include not only Lorentz transformations bulk also translations. The string action is required to be invariant under both of these:
\begin{subequations}\label{eqn:poincare-symmetries}
	\begin{align}
		X^\mu & \rarrow X^\mu + \delta X^\mu; \quad \delta X^\mu \in \mbb{R}^{3,1} \\
		X^\mu & \rarrow \Lambda^\mu{}_\nu X^\mu; \quad \Lambda^\mu{}_\nu \in \mathrm{SO}(3,1)
	\end{align}
\end{subequations}
where $\delta X^\mu$ is some arbitrary vector in Minkowski space. The semidirect product of $\mbb{R}^{3,1}$ with $\mathrm{SO}(3,1)$ is $\mbb{P}^{3,1}$, the Poincaré group in $3+1$ dimensions. Of course, these are good symmetries only if our background is flat. An arbitrarily curved background can have one or more Killing vectors depending on how much symmetry the geometry has, but \emph{in general} even that need not be the case. The best one can hope for is some notion of local flatness is some finite subregion of the background manifold. Moreoever the entire purpose of this work to try and understand how to write down a string action in an \emph{arbitrarily curved} background, assuming that such a thing is even possible.

Even in an arbitrarily curved background, however, symmetry still plays a very important role. Even if we have discarded \emph{global} symmetries, that still leaves us with \emph{local} symmetries in the form of diffeomorphisms under which the action must remain invariant. If, in addition, we have fields such as a gauge connection and an associated vielbein living in the background, then we also have to take local gauge transformations of these objects into account.

However, diffeomorphisms and gauge transformations are not independent of each other. A diffeomorphism is a differentiable map from one manifold to another (in the case of general relativity both manifolds are generally taken to be the same), which has a differentiable (and invertible) map. In more pedestrian language, a diffeomorphism is an (invertible) change of co-ordinates on a manifold. A gauge transformation is a redefinition of the local fields which live on the underlying manifold. In fact, at least for certain geometries, these two can be seen to be equivalent to each other, with diffeomorphisms representing ``passive'' co-ordinate transformations and gauge representing ``active'' co-ordinate transformations.

We will assume henceforth that the Poincaré group is the gauge group which must be promoted from being a \emph{global} symmetry to being a \emph{local} symmetry of the background spacetime.
